Fitting the Undetected: Light-curve Feature Estimation in the Presence of Non-detections

### 2013-03-28 tex/ms.tex
lots of figures added to manuscript

Telescopes don't make catalogs, they make images (Hogg \& Lang 20xx).
Nonetheless, some of the most important astronomical projects have produced enormously productive catalogs
(the \textsl{SDSS}, for examle CITATION).
There is a growing understanding that precise measurements of astronomical sources have to be made directly in the imaging data
(CITE Lang, Anderson, something in weak lensing):
It is impossible to pass the full information content of the imaging to the end user via a catalog
(which by necessity is built with point estimates of the quantities of interest).

The measurements of importance for variable-star science are relatively simple.
Unlike the measurements used for astrometry, extragalactic transients, and weak lensing,
where two-dimensional (morphological) information matters,
when variable stars are being observed, all the end user wants is accurate photometry of each source at each epoch.
One could hope that catalog-level information is sufficient:
the detailed image morphology is not required for lightcurve extraction, except when fields are crowded.
However, there are still limitations to working at catalog level:
Catalog entries (nowadays) usually come with noise variance estimates,
but they don't usually give detailed information about the locations or influence of
bad detector pixels, cosmic rays, and flat-field variations;
all these things might be needed to construct a precise photometric noise model or to understand residuals or artifacts.
More importantly, catalogs are usually subject to hard \emph{censoring};
the catalogs (nowadays) only include measurements that are deemed (by the software or catalog makers)
to be \emph{signficant detections}.

This could change.
In principle a catalog could contain a measurement at \emph{every} epoch for any source that is detected at \emph{any} epoch.
This kind of ``forced photometry'' would include many non-significant measurements,
but it would come closer to passing forward all the information in the imaging to users working at catalog level.
We recommend this approach to the catalog makers of the future (especially those making the \textsl{LSST} catalogs).

Less good than the comprehensive forced-photometry approach is to report, when measurements are not significant,
a well-defined ``upper limit'' for the brightness of each source.
This is rarely done, and upper limits are hard to use, but upper limits are an improvement over standard practice.
Standard practice is to \emph{report nothing at all} when a source is not detected significantly.
Indeed, it is often hard, when presented with a multi-epoch catalog,
to determine whether the reason a particular source is missing from the imaging at a particular epoch
is because the source was faint or because the source fell outside the imaging field of view!

In this contribution we look at improving catalog-level time-domain astronomy in the presence of \emph{censored data}.
The problem we consider is that of period and amplitude modeling for large-amplitude variable stars in a data set
in which \emph{no information is reported at all} at the epochs at which the star is not significantly detected.
Furthermore, we will imagine that we also have \emph{no information at all} about the sensitivity of the telescope
at the time of each non-detection; that is, we have no upper limits and we can't reconstruct them from any extant meta-data.
The non-detections are literally database ``null'' values.
This set-up may seem artificial, but it is in fact the situation we find ourselves in for various surveys,
including the ASAS survey, data from which we will model below.
What we will show is that \emph{the ``null'' non-detections are in fact very informative};
amplitude and period fitting will be wrong in general if the non-detections are ignored.

The benefit of using the non-detections is large, but the cost is also large:
Use of these ``null'' data involves non-trivial modeling,
and large numbers of latent parameters to be inferred and marginalized out.

To make this all concrete, consider the photometric data from the \textsl{ASAS} survey
depicted in \figurename~\ref{fig:example} for one Mira variable.
The photometric detections are shown and so are the epochs at which there are nulls (or non-detections).
The standard

...as an example, consider \figurename~\ref{fig:example}...

@@

\clearpage
\begin{figure}
  \begin{center}
\includegraphics[angle=0,width=6.5in]{../plots/fig_045832-0604_1.png}
  \end{center}
  \caption{A comparison of the best fit obtained by the traditional method (left) and our censored data model (right) on the Mira variable ASAS 045832-0604.1.  The top row shows the respective model fits to the raw light curves while the bottom row displays the period-folded data and models.  The black dots denote the observed data and the red vertical dashes mark the epochs of null observation.  The traditional fitting method clearly obtains half the true period while our method recovers the correct period.  The traditional method fails because it identifies a reasonable high-frequency fit to the observed data which completely ignores the null observations.  Note that alternate cycles of the traditional model fit contain no observed data.  By contrast, our method fully accounts for the null observations using a probability model.\label{fig:badfit}}
\end{figure}

@@

\clearpage
\begin{figure}
\begin{center}
\includegraphics[angle=0,width=6.5in]{../plots/mira_simulated.pdf}
\end{center}
\caption{ Folded light curves for data simulated from the Mira variable ASAS 235627-4947.2 using the prescription in \S\ref{ss:mirasim}.  Blue tick marks along the bottom axis denote phases where there were non-detections.  As the original flux measurements are dimmed by higher factors, the troughs of the sinusoidal light curve are censored, resulting in an incomplete view of the data.  \hogg{This figure needs to show the OLD and NEW best-fit lightcurve models too!}\label{fig:mirasim}}
\end{figure}

\clearpage
\begin{figure}
  \begin{center}
    [\hogg{PUT FIGURE HERE}]
  \end{center}
  \caption{A real-data example.  This is the same source as that shown in \figurename~\ref{fig:badfit} but now showing the results of...\label{fig:goodfit}}
\end{figure}

\clearpage
\begin{figure}
  \begin{center}
    [\hogg{PUT FIGURE HERE}]
  \end{center}
  \caption{Another few real-data examples; in each case the method presented here either matches the traditional method or is substantially better.\label{fig:examples}}
\end{figure}

>>>>

Surveys working in real-time tend to analyze each image with an eye to populating databases quickly. Unless a particular region of the sky gets special treatment (such as it is a known place of interest), discovery of variability happens by comparison of flux values at the database level. Analysis on such databases of transient/variables thus deals with censored data where there is often less information available than there should be. Even then, we might be able to infer what the thresholds were based on S/N measurements of objects neighboring the source of interest.

@@

\clearpage
\begin{figure}
  \begin{center}
\includegraphics[angle=0,width=6.5in]{fig_045832-0604_1.png}
  \end{center}
  \caption{A comparison of the best fit obtained by the traditional method (left) and our censored data model (right) on the Mira variable ASAS 045832-0604.1.  The top row shows the respective model fits to the raw light curves while the bottom row displays the period-folded data and models.  The black dots denote the observed data and the red vertical dashes mark the epochs of null observation.  The traditional fitting method clearly obtains half the true period while our method recovers the correct period.  The traditional method fails because it identifies a reasonable high-frequency fit to the observed data which completely ignores the null observations.  Note that alternate cycles of the traditional model fit contain no observed data.  By contrast, our method fully accounts for the null observations using a probability model.\label{fig:badfit}}
\end{figure}

@@

\clearpage
\begin{figure}
\begin{center}
$\begin{array}{c@{\hspace{-.3in}}c}
\multicolumn{1}{l}{\mbox{\bf (a)}} & \multicolumn{1}{l}{\mbox{\bf (b)}} \\      \vspace{-.1in}
\includegraphics[angle=0,width=3.5in]{fig_simulation_0_01_dat.png} & \includegraphics[angle=0,width=3.5in]{fig_simulation_0_02_dat.png} \\
\multicolumn{1}{l}{\mbox{\bf (c)}} & \multicolumn{1}{l}{\mbox{\bf (d)}} \\
\includegraphics[angle=0,width=3.5in]{fig_simulation_0_05_dat.png} & \includegraphics[angle=0,width=3.5in]{fig_simulation_0_1_dat.png}
\end{array}$
%\includegraphics[angle=0,width=6.5in]{../plots/mira_simulated.pdf}
\end{center}
\caption{ \joey{rewrite this caption!} Folded light curves for data simulated from the Mira variable ASAS 235627-4947.2 using the prescription in \S\ref{ss:mirasim}.  Blue tick marks along the bottom axis denote phases where there were non-detections.  As the original flux measurements are dimmed by higher factors, the troughs of the sinusoidal light curve are censored, resulting in an incomplete view of the data.  \hogg{This figure needs to show the OLD and NEW best-fit lightcurve models too!}\label{fig:mirasim}}
\end{figure}

\clearpage
\begin{figure}
  \begin{center}
$\begin{array}{c@{\hspace{-.3in}}c}
\multicolumn{1}{l}{\mbox{\bf (a)} \hspace{0.7in} \textrm{ ASAS 034344+0655.5}} & \multicolumn{1}{l}{\mbox{\bf (b)}\hspace{0.7in}  \textrm{ ASAS 091617-2936.7}} \\      \vspace{-.1in}
\includegraphics[angle=0,width=3.5in]{fig_034344+0655_5.png} & \includegraphics[angle=0,width=3.5in]{fig_091617-2936_7.png} \\
\multicolumn{1}{l}{\mbox{\bf (c)}\hspace{0.7in}  \textrm{ ASAS 075855-1914.1}} & \multicolumn{1}{l}{\mbox{\bf (d)}\hspace{0.7in}  \textrm{ ASAS 075826-4019.8}} \\
\includegraphics[angle=0,width=3.5in]{fig_075855-1914_1.png} & \includegraphics[angle=0,width=3.5in]{fig_075826-4019_8.png}
\end{array}$
  \end{center}
  \caption{ \joey{add caption: period doubling} A real-data example.  This is the same source as that shown in \figurename~\ref{fig:badfit} but now showing the results of...\label{fig:goodfit}}
\end{figure}

\clearpage
\begin{figure}
  \begin{center}
$\begin{array}{c@{\hspace{-.3in}}c}
\multicolumn{1}{l}{\mbox{\bf (a)} \hspace{0.7in} \textrm{ ASAS 112811-6606.3}} & \multicolumn{1}{l}{\mbox{\bf (b)}\hspace{0.7in}  \textrm{ ASAS 065002-4554.6}} \\      \vspace{-.1in}
\includegraphics[angle=0,width=3.5in]{fig_112811-6606_3.png} & \includegraphics[angle=0,width=3.5in]{fig_065002-4554_6.png} \\
\multicolumn{1}{l}{\mbox{\bf (c)}\hspace{0.7in}  \textrm{ ASAS 092021-5815.2}} & \multicolumn{1}{l}{\mbox{\bf (d)}\hspace{0.7in}  \textrm{ ASAS 082012-3816.7}} \\
\includegraphics[angle=0,width=3.5in]{fig_092021-5815_2.png} & \includegraphics[angle=0,width=3.5in]{fig_082012-3816_7.png}
\end{array}$
  \end{center}
  \caption{\joey{add caption: period tripling} Another few real-data examples; in each case the method presented here either matches the traditional method or is substantially better.\label{fig:examples}}
\end{figure}
